---
title: "Reproducible Research: Peer Assessment 2"
author: "GitHub: atet"
date: "July 14, 2014"
output: html_document
---

---

# TITLE:

---

## Synopsis

NULL

---

## Data Processing
A GitHub repository was created for this document at: https://github.com/atet/coursera/tree/master/RepData_PeerAssessment2

### Loading Data
  * The original required data file archive `repdata-data-StormData.csv.bz2` and the prepared compressed `repdata-data-StormData.rds` is included in the repository
  * The following code is to convert the original `repdata-data-StormData.csv.bz2` to a compressed R object `repdata-data-StormData.rds`. Since the archive uncompressed is > 100 MB, the following code will download the archive from online, uncompress it locally to home (`~/`), load the local `repdata-data-StormData.csv` file into an R object, and then save it out as `repdata-data-StormData.rds` (which is below the GitHub 100 MB limit):
```{r load_bz2, eval=FALSE}
setwd("~/")
filepath_source      <- "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2"
filename             <- "repdata-data-StormData"
filepath_destination <- paste("~/", filename, sep = "")
download.file(url = filepath_source, destfile = paste(filepath_destination, ".csv.bz2", sep = ""), method = "curl")
data                 <- bzfile(description = paste(filepath_destination, ".csv.bz2", sep = ""))
data                 <- read.csv(data, stringsAsFactors = FALSE) # Remember to set stringsAsFactors = FALSE
saveRDS(data, "~/repdata-data-StormData.rds")
# NOTE: repdata-data-StormData.rds is then uploaded to my GitHub repository
# We will use this for the remained of the analysis
```
  * This following code will now load the compressed `repdata-data-StormData.rds` from GitHub into R.
```{r load_rds}
setwd("~/")
filepath_source      <- "https://raw.github.com/atet/coursera/blob/master/RepData_PeerAssessment2/repdata-data-StormData.rds"
filename             <- "repdata-data-StormData"
filepath_destination <- paste("~/", filename, sep = "")
download.file(url = filepath_source, destfile = paste(filepath_destination, ".rds", sep = ""), method = "curl")
data                 <- readRDS(paste(filepath_destination, ".rds", sep = ""))
```


---

## Results

NULL

---

## Discussion

### 1. Across the United States, which types of events (as indicated in the EVTYPE variable) are most harmful with respect to population health?

### 2. Across the United States, which types of events have the greatest economic consequences?

---
---